Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-12 manuscript.rmd
Added intro para on realised water-environments

```{r, include=FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
```

```{r setup}
library(knitr)
opts_knit$set(eval.after = "fig.cap")
```

\
**Corresponding author**: 
Rutuja Chitra-Tarak
\
E: arutuj@gmail.com
\
T: +1(505)

@@

Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome based on current estimates of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

@@

\
Here, we obtained daily dynamics of plant water potential ($\Psi_{soil, z}$) in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model. We parametarised this model using local soil properties such as hydraulic conductivity by depth and water retention curves, and calibrated it against new data on soil moisture dynamics by depth, as well as catchment evapotranspiration and stream discharge. We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{soil, z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model. This model incorporated impact of atmospheric drought (VPD) on growth as well as new data on species-specific leaf hydraulic conductivity responses to water stress ($K_{Leaf}$ vs. $\Psi_{x}$). As vulnerability curves were collected for a small fraction of species (hard trait), we developed a new model that predicted the parameters of these curves from Leaf Mass Area and Wood Specific Gravity i.e. leveraging soft traits available community-wide. We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).

We test three hypotheses under an overarching one that interspecific variation in effective rooting depths of trees aligns with a conservative-acquisitive trade-off axis.  In this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even through multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) canpoy trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

@@

\
We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode (0.1$^\circ$ x 0.1$^\circ$ centered at 9.1530 $^\circ$N, 280.2539 $^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter waws chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

| S.N. | Parameter | Description | Range (Units) | Method of determination with reference
| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- |
| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of (Medlyn et al. 2011) | 
| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2/gC$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $g/m^2$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass. |  
| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at 25 $^\circ$C, canopy top | 22.7 - 92.5 umol $CO_2/m^2/s$ | Observed range of Vcmax at 25 $^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves. Values derived with a conversion specific for CLM model are used. (Ali et al. 2015) | 
| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq. 2 in (Zeng 2001) that regulates the shape of the rooting profile. Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 (or 2) of (Zeng 2001). | 
| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq. 2 in (Zeng 2001) that regulates the depth of the rooting profile. Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined (Canadell et al. 1996). Besides the direct observation of roots at 18 m included by (Nepstad et al. 1994) in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m. (Davidson et al. Forest Science 2011) |
| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species. This study| 
| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mm/s for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm | To account for high macroporosity and direct flow paths in tropical soils(Broedel et al. 2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al. 2017; Tomasella and Hodnett 1996) For Conrad catchment observed median Ksat (95%CI) for 12.5 cm depth is 38.3 mm/hr (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mm/hr. For reference a storm with 12.5 mm/hr rainfall intensity has a 0.2 probability of occuring in any given rainfall event (Kinner and Stallard 2004). | 
| 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples (Broedel et al. 2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al. 2017; Tomasella and Hodnett 1996) | 
| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05-0.44 unitless | Based on throughfall data from (Zimmermann et al. 2010) and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm. |

### Appendix A

#### 1. 
# References

>>>>

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
```

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(eval.after = "fig.cap")
```

\
**Corresponding author**: 
Rutuja Chitra-Tarak
\
E: arutuj@gmail.com
\
T: +1(505) 665 1647

@@

Keywords: rooting depths, drought tolerance, mortality, tropical forest, climate change; life-history strategies

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome based on observed ranges of trees' hydraulic safety margins [@Choat:2012; but see @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

@@

Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be established for trees' realised water environments and water-access through droughts. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88, Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88, Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88, Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88, Stem}$). A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88, Stem}$ than other two species limited to shallow-water. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. Authors concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated with rooting/water-sourcing depths, or related to hydraulic traits.

\
Here, we obtained daily dynamics of plant water potential ($\Psi_{soil, z}$) in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model. We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{soil, z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model. This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x}$), which we scaled-up to community level via scaling relationships with commonly available traits. We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).

We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns them on a conservative-acquisitive trade-off axis. Specifically, in this forest (i) dynamics of soil water potential at depth does not cross catastrophic levels even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio; (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

@@

\
We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter waws chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

| S.N. | Parameter | Description | Range (Units) | Method of determination with reference
| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- |
| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of (Medlyn et al. 2011) | 
| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass. |  
| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of Vcmax at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves. Values derived with a conversion specific for CLM model are used. (Ali et al. 2015) | 
| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq. 2 in (Zeng 2001) that regulates the shape of the rooting profile. Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 (or 2) of (Zeng 2001). | 
| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq. 2 in (Zeng 2001) that regulates the depth of the rooting profile. Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined (Canadell et al. 1996). Besides the direct observation of roots at 18 m included by (Nepstad et al. 1994) in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m. (Davidson et al. Forest Science 2011) |
| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species. This study| 
| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
| 9 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mms{-1} for $\le$ 12.5 cm; 5.56e-05, 0.0003 for $\ge$ 60 cm | To account for high macroporosity and direct flow paths in tropical soils(Broedel et al. 2017; Tomasella and Hodnett 1996) (Kinner and Stallard 2004; Broedel et al. 2017; Tomasella and Hodnett 1996) For Conrad catchment observed median Ksat (95% CI) for 12.5 cm depth is 38.3 mm/hr (25.4-51.2, *n* = 75), while for 60 cm depth 0.7 (0.2, 1.2, *n* = 40) mmhr^{-1}^. For reference a storm with 12.5 mmhr^{-1}^ rainfall intensity has a 0.2 probability of occuring in any given rainfall event (Kinner and Stallard 2004). | 
| 10 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples (Broedel et al. 2017; Tomasella and Hodnett 1996)(Kinner and Stallard 2004; Broedel et al. 2017; Tomasella and Hodnett 1996) | 
| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from (Zimmermann et al. 2010) and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm. |

# References
